NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-01-05 manuscriptPGPE.tex

  %The force field giving the best description for glycerol backbone and headgroup structures
  %of PC, PS, PG and PE headgroups (CHARMM36) reproduces the essential differences in order parameters
  %between these headgroups, and therefore enables the analysis of structural differences between the
  %headgroups.

@@

Lipid conformational ensembles in physiologically relevant lamellar liquid phase
are typically derived from NMR experiments, particularly from 
%Attempts to resolve the conformational ensembles of lipid headgroup rely mainly on
C-H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}
which can be detected even from cells \cite{gally81,scherer87,seelig90}.
%but lipids are not in solid state in biomembranes. 
%On the other hand, NMR experiments enable detection of lipid headgroups
%in physiological liquid states also in cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone structures are largely similar irrespectively of the headgroup \cite{gally81}, 
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
%Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
%but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
However, these results are based only on the absolute values while 
%we have recently demonstrated that also
the necessity of order parameters signs in capturing the lipid conformational ensembles has been resently demonstrated \cite{botan15,ollila16,ferreira16}.
Furthermore, the detailed understanding of lipid conformational ensembles
is limited by the lack of universal models to interpret the experiments \cite{pezeshkian18,akutsu20}.
%do not exist and
%the proposed structures are often based on lipid crystal structures \cite{pezeshkian18,akutsu20}.
%which were solved many decades ago \cite{buldt81,pascher92}.
%determine the structural 
%Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
%the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
%force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.

Stuctures of different lipid types in protein bound states can be extracted from the protein data bank (PDB) \cite{berman00},
but their relation to the conformational ensembles in liquid lamellar state remains unclear \cite{marsh13b}.
In addition to the changes in lipid conformational ensembles upon binding to proteins,
%PC headgroup conformations do not change upon addition of zwitterionic lipids or cholesterol \cite{scherer87,ferreira13} into a membrane,
%but are sensitive to the presence of charged molecules in membranes \cite{seelig87}.
also the experimentally measured response of lipid headgroup to membrane bound charges remains poorly understood 
%The response to charges is explained by the change in the orientation of lipid headgroup dipole vector due to electrostatic interactions,
%but the detailed changes are not yet understood
due to the lack of suitable models to interpret the lipid conformational ensembles in liquid lamellar state \cite{Semchyschyn04}.

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
%open collaboration and order parameters of glycerol backbone and headgroup
%to evaluate the accuracy of PE and PG heagroup structures, and the cation binding affinity
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
%PE and PG lipids are most common lipids in bacteria \cite{sohlenkamp16}.
%Anionic PG lipids are less abundant, but is also proposed to be fundamental for terrestrial life \cite{furse17}.
%PE and PG affect membrane protein functionality \cite{hariharan18} and
All the studied lipids specifically bind to various proteins \cite{yeagle14}.
%PE headgroup is also prone for negative membrane curvature and causes membrane fusion \cite{Chernomordik08,calzada16}.
%Therefore, the PE and PG headgroup structures play probably essential roles in 
%many biological processes.
Furthermore, we use our results to elucidate lipid-protein interactions and the effect of charges on lipid conformations.

@@

%Several MD simulations of PE and PG lipids have been published especially in the context of modeling
%inner membrane of Gram-negative
%bacteria \cite{devries04,murzyn05,pedersen06,zhao07,gurtovenko08,zhao08,henin09,kukol09,tsai12,dickson12,venable13,dickson14,berglund15}
%\todo{There may be some relevant publication missing from here},
%but evaluation of glycerol backbone and headgroup structures against experiments is rare \cite{henin09}.

%Derivation of different lipid headgroup conformational ensembles in liquid state has been
%inconclusive due to lack of suitable experimental data and tools to interpret the conformational ensembles.

%Besides the structure, also ion binding may regulate biophysical activity of especially negatively
%charged lipid headgroups \cite{seelig90}. Monovalent cation (except Lithium) binding to zwitterionic PC and anionic PS headgroups is
%very weak, while multivalent ion binding is stronger but still weak \cite{cevc90,tocanne90,roux90,catte16,antila19}. The ion binding affinity
%data for PE is more scarce \cite{marra85}, but large differences to PC would be surprising.
%Negatively charged lipids are suggested to bear same cation binding constants than zwitterionic lipids, but
%the amount of bound ions to negatively charged membranes would still be larger because
%the concentration of cations in the vicinity of membranes would be higher  \cite{seelig90}.
%
%is stronger
%Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}
%
%On the other hand, anionic PS lipids are proposed chelate with calcium ions \cite{hauser85,feigenson86,roux91}.
%In simulations, the cation binding affinity to PC and PS membranes is typically overestimated \cite{catte16,antila19},
%which can be improved by applying the ECC to the partial charges of the force fields \cite{melcr18,melcr19}.

%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
%  \caption{\label{lipids}
%    Chemical structures and labels for the headgroup carbons.
%  }
%\end{figure}

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of POPE and POPG
were measured using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings
using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
The NMR experiments were identical as in our previous work~\cite{antila19}.
POPE and POPG powder were purchased from Avanti polar lipids \todo{Which isomer of POPG we have in the experiments?}.
%\todo{Is this enough and correct, or should we repeat some methods from the NMRlipidsIVps paper?}
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

%Absolute values of the headgroup and glycerol backbone order parameters from PE and PG lipids are measured
%previously using $^2$H NMR~\cite{seelig76,gally81,wohlgemuth80,borle85}. Because also the order parameter
%signs bear essential information about the lipid structures \cite{botan15,ollila16}, we measured the
%magnitudes and signs of POPE and POPG C--H bond headgroup and glycerol backbone order parameter in liquid phase
%using the 2D-RPDLF and S-DROSS experiments, as described previously \cite{ferreira13,ferreira16,antila19}.

Glycerol backbone peaks from both lipids, and $\alpha$-carbon peak from POPE in the INEPT spectra 
were assigned based on previously measured POPC spectra~\cite{ferreira13}.
%For POPE, the glycerol backbone and $\alpha$-carbon peaks in INEPT spectra were assigned based on
%previously measured POPC spectra~\cite{ferreira13} and
The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
%For POPG, the glycerol backbone peaks in INEPT spectra were assigned based on
%previously measured POPC spectra~\cite{ferreira13},
\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
because their chemical environments are similar in POPG.
\todo{Details to be checked by Tiago}.

@@

%\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}

\begin{figure*}[]
  \centering
   \includegraphics[width=18.0cm]{../Figs/figure2.eps}
%  \includegraphics[width=18.0cm]{../Figs/PCPEPGPSstructures2.png}
%  \includegraphics[width=8.0cm]{../Figs/CHARMMfromLIPIDS.eps}
%%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals1.png}
%%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals2.png}
%  \includegraphics[width=8.0cm]{../Figs/PNangleNORM.eps}
%  \includegraphics[width=8.0cm]{../Figs/DIHEDRALS.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles betwee different lipids. 
     {\bf A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     {\bf B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     {\bf C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     {\bf D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids
that are not bound to proteins in electrostatically neutral cell membrane,
we measured the C-H bond order parameters and their signs from of POPG and POPE
in liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the
%R-PDLF and S-DROSS
data in Figs. \ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C-H bonds, except the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig. \ref{HGorderParameters} B).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
%order parameter of $\beta$-carbon is expected to be clearly smaller than for 
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig. \ref{HGorderParameters} (B)).
This interpretation is confirmed by SIMPSON calculations in Fig \ref{POPGsimpson}.
%using negative value for g$_2$ and positive value for $\beta$ order parameter (Fig. \ref{POPGsimpson}).
%The signs of these order parameters were solved using SIMPSON simulations to interpre the
%S-DROSS experiments in similar fashion as we did previously for PS lipids \cite{antila19}.

%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
%While the headgroup order parameters are similar for PC and PE lipids,
%PG and PS lipids exhibit distinct values in comparison with other lipids.

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and
headgroups from this and previous studies are collected in Fig. \ref{HGorderParameters} D),
where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
%The resulting order parameters are compared with our previously published results
%for POPC \cite{ferreira13,ferreira16} and POPS \cite{antila19} in figure \ref{HGorderParameters} D.
%Order parameters for POPE are in good agreement with previous $^2$H NMR experiments \cite{seelig76}
%and similar to our previous results for POPC \cite{ferreira16}.
Overall agreement of order parameters determined by different authors and different techniques
for same lipid headgroup is very good in here and previous studies \cite{botan15,ollila16,antila19},
suggesting that the observed differences between lipid types arise from differences in
headgroup chemistry rather than inaccuracies in experiments. 
The most distinct order parameters are observed for PS headgroups,
for which the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
On the other hand, the $\beta$-carbon order parameter of PG headgroup
%the $\alpha$-carbon order parameter is similar to PE and PG, while
has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
%because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
Essential differences between PC and PE headgroups are not observed.

%As discussed previously for PC and PS headgroups \cite{ollila16,antila19}, also 
%the headgroup and glycerol backbone order parameters of PE and PG are essentially independent of 
%acyl chain composition, and therefore manifest mainly headgroup chemistry (Fig. \ref{HGorderParameters}).

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{../Figs/HGorderparametersPCvsSURFchangeDIHEDRALS.eps}
  \caption{\label{changesWITHsurf}
    {\bf A)} Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with experimental data \cite{??}.
    {\bf B)} Changes in PC headgroup conformational ensembles upon increasing amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

To resolve the conformational ensembles of lipid headgroups,
we compared the headgroup and glycerol backbone C-H bond order parameters
from different MD simulation force fields to the experimental data. 
None of the force fields reproduced the order parameters
within experimental accuracy for any of the lipids analysed in this work
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}.
%has least problems in
%reproducing the experimental lipid headgroup and glycerol backbone order parameters in different
%lipids \cite{botan15,antila19} (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
Nevertheless, the CHARMM36 force field performs best for all the headgroups and
%Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).
%To characterize the differences in lipid headgroup conformational ensembles
reproduces the distinct order parameters of PS and PG lipids (Fig. \ref{structures} B).
To understand the structural origin of these distinct order parameters,
we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in figure \ref{structures} C).
%The dihedral distributions in figure \ref{structures} show significant similarity between different lipids for all the other
%bonds, except
Major differences between headgroups are observed only for the last two dihedrals
in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$,
which prefer {\it gauche$^-$} conformations for PG and {\it gauche$^+$} for PS.
Rest of the dihedrals are similar between different lipids, with the exception of
%However, some dihedral angles of phosphorus-oxygen bonds are less likely for PS, which
small differences for PS lipids. Therefore we suggest that the main differences between
lipid headgroups leading to distinct order parameter occur in the choline part, while
also changes in phosphate region may contribute in PS lipids, which could also
explain the more rigid headgroup structures~\cite{browning80,buldt81}.
%These differences explaing the disctinct order parameters of
%PS and PG headgroups with respenct to the other lipids.
%The differences between lipids are reflected also to
Furthermore, the angle between headgroup dipole and membrane normal
decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures} B)).
However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P-N vector dipole may be an artefacts as the $\beta$-carbon order parameter in PC
is too negative the CHARMM36 force field, thereby not being equal to the order parameter
in PE as observed in experiments \cite{botan15}.

%The eclipsed (0 or 360 degrees) are not present in any other dihedrals, and anti-eclipsed (120 or 240 degrees)
%are not present in C$_\alpha$-C$_\beta$ and g$_1$-g$_2$ bonds.

%Characteristic dihedral conformations in PS headgroup are asymmetric conformations
%preferring gauche 270$^\circ$ conformations in N-C$_\beta$-C$_\alpha$-O$_\alpha$ and
%%dihedral exhibits a more asymmetric angle distribution for PS, having largest probability being
%%in , than for the PC headgroup.
%C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedrals. 
%In PG headgroup, the O$_\beta$-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral (corresponding N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral in other% lipids)
%is mostly in trans conformation, and C$_\beta$-C$_\alpha$-O$_\alpha$-P  has asymmetric tendency to be in
%gauche 60$^\circ$ conformation. 
%Main difference between PC and PE is the lower probability of trans state in C$_\beta$-C$_\alpha$-O$_\alpha$-P PC dihedral,
%which could be a potential reason for the too negative $\beta$ headgroup order parameter in PC.%

In conclusion, all lipid headgroups sample very broad conformational ensembles in liquid lamellar phase.
%the results suggest that free rotations around O-P bonds decouple the
%headgroup and glycerol backbone structure and dynamics in similarly in all lipids,
%althought this rotation is slower in PS lipids where crossing certain barriers are slower.
Despite the differences in distributions
%of two the last bonds at the headgroup end,
%leading to the distinct order parameters for PS and PG lipids,
%lipids with
between headgroups, the sampled dihedral angles are within approximately same ranges for all headgroup types.
Therefore, we suggest that all lipid headgroups are very flexible thereby being
able to adopt wide range of multiple conformations when interacting with proteins, ions or other biomolecules.
%Furthermore, approximately same conformations are present liquid lamellar phase for different
%headgroups studied here, althought the probability weight for the conformations are slightly different.
%the order parameter experiments suggest that the glycerol backbone conformations in all lipids
%and the headgroup conformations in PC and PE lipids are relatively similar, while 
%PS and PG headgropus exhibit distinct conformational sampling.
%The details of sampled conformation are difficult to deduce from order parameters only,
%but the distinct headgroup order parameters
%of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.
%The analysis of dihedral angle distributions reveal the free rotation of dihedral around
%phosphorus-oxygen bonds in all lipids as all possible angles are observed in the distributions.
Our results support the models that propose free rotation around phosphate group for PC lipids,
which decouples the dynamics and conformatios between acyl chains and headgroups \cite{klauda08c,antila21b},
and suggests that these models could be applied for all lipids containing the phosphorus group.
The wide range of observed conformations suggest that the structures in lipid crystals \cite{buldt81,pascher92}
play only a minor role, and that the models aiming to explain NMR data using only few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20}
are not sufficient to capture the large conformational space of lipids in liquid lamellar state.

@@

Charged lipids, proteins, surfactants, drugs, and ions incorporated in membranes
reorient the headgroup dipole in PC lipids, thereby affecting the order parameters of
lipid headgroups \cite{seelig87}. However, detailed understanding on
lipid conformational ensembles in membranes bearing electric charge is still lacking \cite{Semchyschyn04}.
%Order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup are known to depend
%linearly on the incorporation of charged molecules into a lipid bilayer,
%which is explained by tilting of the headgroup dipole orientation \cite{??}.
%%are kn affect the headgroup dipole vector tilt which can
%%be detected by measuring C-H bond order parameters of $\alpha$ and $\beta$ carbons.
%Such changes have been observed upon addition of charged lipids, proteins, surfactants,
%drugs, or ions \cite{??}, but how charges affect the lipid conformational ensemble remains unknown.

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    Calsium density distributions are shown in figure \ref{CAdensPG}.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

>>>>

@@

Lipid conformational ensembles in physiologically relevant lamellar liquid phase
are typically derived from NMR experiments, particularly from 
C-H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}
which can be detected even from cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone structures are largely similar irrespectively of the headgroup \cite{gally81}, 
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
However, these results are based only on the absolute values while 
the necessity of order parameters signs in capturing the lipid conformational ensembles has been resently demonstrated \cite{botan15,ollila16,ferreira16}.
Furthermore, the detailed understanding of lipid conformational ensembles
is limited by the lack of universal models to interpret the experiments \cite{pezeshkian18,akutsu20}.

Stuctures of different lipid types in protein bound states can be extracted from the protein data bank (PDB) \cite{berman00},
but their relation to the conformational ensembles in liquid lamellar state remains unclear \cite{marsh13b}.
In addition to the changes in lipid conformational ensembles upon binding to proteins,
also the experimentally measured response of lipid headgroup to membrane bound charges remains poorly understood 
due to the lack of suitable models to interpret the lipid conformational ensembles in liquid lamellar state \cite{Semchyschyn04}.

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the studied lipids specifically bind to various proteins \cite{yeagle14}.
Furthermore, we use our results to elucidate lipid-protein interactions and the effect of charges on lipid conformations.

@@

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of POPE and POPG
were measured using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings
using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The NMR experiments were identical as in our previous work~\cite{antila19}.
POPE and POPG powder were purchased from Avanti polar lipids \todo{Which isomer of POPG we have in the experiments?}.
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

Glycerol backbone peaks from both lipids, and $\alpha$-carbon peak from POPE in the INEPT spectra 
were assigned based on previously measured POPC spectra~\cite{ferreira13}.
The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
because their chemical environments are similar in POPG.
\todo{Details to be checked by Tiago}.

@@

\begin{figure*}[]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles betwee different lipids. 
     {\bf A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     {\bf B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     {\bf C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     {\bf D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids
that are not bound to proteins in electrostatically neutral cell membrane,
we measured the C-H bond order parameters and their signs from of POPG and POPE
in liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the data in Figs. \ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C-H bonds, except the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig. \ref{HGorderParameters} B).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig. \ref{HGorderParameters} (B)).
This interpretation is confirmed by SIMPSON calculations in Fig \ref{POPGsimpson}.

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and
headgroups from this and previous studies are collected in Fig. \ref{HGorderParameters} D),
where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature. Overall agreement of order parameters determined by different authors and different techniques
for same lipid headgroup is very good in here and previous studies \cite{botan15,ollila16,antila19},
suggesting that the observed differences between lipid types arise from differences in
headgroup chemistry rather than inaccuracies in experiments. 
The most distinct order parameters are observed for PS headgroups,
for which the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
On the other hand, the $\beta$-carbon order parameter of PG headgroup
has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
Essential differences between PC and PE headgroups are not observed.

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALS.eps}
  \caption{\label{changesWITHsurf}
    {\bf A)} Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with experimental data \cite{??}.
    {\bf B)} Changes in PC headgroup conformational ensembles upon increasing amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

To resolve the conformational ensembles of lipid headgroups,
we compared the headgroup and glycerol backbone C-H bond order parameters
from different MD simulation force fields to the experimental data. 
None of the force fields reproduced the order parameters
within experimental accuracy for any of the lipids analysed in this work
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}.
Nevertheless, the CHARMM36 force field performs best for all the headgroups and
reproduces the distinct order parameters of PS and PG lipids (Fig. \ref{structures} B).
To understand the structural origin of these distinct order parameters,
we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in figure \ref{structures} C).
Major differences between headgroups are observed only for the last two dihedrals
in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$,
which prefer {\it gauche$^-$} conformations for PG and {\it gauche$^+$} for PS.
Rest of the dihedrals are similar between different lipids, with the exception of
small differences for PS lipids. Therefore we suggest that the main differences between
lipid headgroups leading to distinct order parameter occur in the choline part, while
also changes in phosphate region may contribute in PS lipids, which could also
explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal
decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures} B)).
However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P-N vector dipole may be an artefacts as the $\beta$-carbon order parameter in PC
is too negative the CHARMM36 force field, thereby not being equal to the order parameter
in PE as observed in experiments \cite{botan15}.

In conclusion, all lipid headgroups sample very broad conformational ensembles in liquid lamellar phase.
between headgroups, the sampled dihedral angles are within approximately same ranges for all headgroup types. Therefore, we suggest that all lipid headgroups are very flexible thereby being
able to adopt wide range of multiple conformations when interacting with proteins, ions or other biomolecules. Our results support the models that propose free rotation around phosphate group for PC lipids, which decouples the dynamics and conformatios between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group. The wide range of observed conformations suggest that the structures in lipid crystals \cite{buldt81,pascher92} play only a minor role, and that the models aiming to explain NMR data using only few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in liquid lamellar state.

@@

Charged lipids, proteins, surfactants, drugs, and ions incorporated in membranes
reorient the headgroup dipole in PC lipids, thereby affecting the order parameters of
lipid headgroups \cite{seelig87}. However, detailed understanding on
lipid conformational ensembles in membranes bearing electric charge is still lacking \cite{Semchyschyn04}.

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{./Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    Calsium density distributions are shown in figure \ref{CAdensPG}.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}
